Mesolimbic dopamine ramps reflect environmental timescales

Version 1

Abstract

Mesolimbic dopamine activity occasionally exhibits ramping dynamics, reigniting debate on theories of dopamine signaling. This debate is ongoing partly because the experimental conditions under which dopamine ramps emerge remain poorly understood. Here, we show that during Pavlovian and instrumental conditioning, mesolimbic dopamine ramps are only observed when the inter-trial interval is short relative to the trial period. These results constrain theories of dopamine signaling and identify a critical variable determining the emergence of dopamine ramps.

Mesolimbic dopamine activity was classically thought to operate in either a “phasic” or a “tonic” mode(1–3). Yet, recent evidence points to a “quasi-phasic” mode in which mesolimbic dopamine activity exhibits ramping dynamics(4– 16). This discovery reignited debate on theories of dopamine function as it appeared inconsistent with the dominant theory that dopamine signaling conveys temporal difference reward prediction error (RPE)(17). Recent work has hypothesized that dopamine ramps reflect the value of ongoing states(1, 4– 6), RPE under some assumptions(8, 10, 18, 19), or a causal influence of actions on rewards in instrumental tasks(11). This debate has been exacerbated in part because there is no clear understanding of why dopamine ramps appear only under some experimental conditions. Accordingly, uncovering a unifying principle of the conditions under which dopamine ramps appear will provide important constraints on theories of dopamine function(1, 11, 18, 20–39).

In our recent work proposing that dopamine acts as a teaching signal for causal learning by representing the Adjusted Net Contingency for Causal Relations (ANCCR)(20, 34, 40), simulated ANCCR depends on the duration of a memory trace of past events (i.e., on the “eligibility trace” time constant) (illustrated in Extended Data Fig 1). Accordingly, we successfully simulated dopamine ramping dynamics assuming two conditions: a dynamic progression of cues that signal temporal proximity to reward, and a small eligibility trace time constant relative to the trial period(20). However, whether these conditions are sufficient to experimentally produce mesolimbic dopamine ramps in vivo remains untested. To test this prediction in Pavlovian conditioning, we measured mesolimbic dopamine release in the nucleus accumbens core using a dopamine sensor (dLight1.3b)(41) in an auditory cuereward task. We varied both the presence or absence of a progression of cues indicating reward proximity (“dynamic” vs “fixed” tone) and the inter-trial interval (ITI) duration (short vs long ITI). Varying the ITI was critical because our theory predicts that the ITI is a variable controlling the eligibility trace time constant, such that a short ITI would produce a small time constant relative to the cue-reward interval (Supplementary Note 1) (Fig 1a-e). In all four task conditions, head-fixed mice learned to anticipate the sucrose reward, as reflected by anticipatory licking (Fig 1f-g). In line with our earlier work, we showed that simulations of ANCCR exhibit a larger cue onset response when the ITI is long and exhibit ramps only when the ITI is short (Fig 1h). Consistent with these simulated predictions, experimentally measured mesolimbic dopamine release had a much higher cue onset response for long ITI (Fig 1i-j). Furthermore, dopamine ramps were observed only when the ITI was short and the tone was dynamic (Fig 1i, k-m). The presence of dopamine ramps during the last five seconds of the cue could not be explained by variations in behavior, as anticipatory licking during this period was similar across all conditions (Extended Data Fig 2, Supplementary Note 2). Indeed, dopamine ramps—quantified by a positive slope of dopamine response vs time within trial over the last five seconds of the cue—appeared on the first day after transition from a long ITI/dynamic tone condition to a short ITI/dynamic tone condition and disappeared on the first day after transition from a short ITI/dynamic tone condition to a short ITI/fixed tone condition (Fig 1l). These results confirm the key prediction of our theory in Pavlovian conditioning.

Pavlovian conditioning dopamine ramps depend on ITI.

a. Top, fiber photometry approach schematic for nucleus accumbens core (NAcC) dLight recordings. Bottom, head-fixed mouse. b. Pavlovian conditioning task setup. Trials consisted of an 8 s auditory cue followed by sucrose reward delivery 1 s later. c. Cumulative Distribution Function (CDF) of ITI duration for long (solid line, mean 55 s) and short ITI (dashed line, mean 8 s) conditions. d. Experimental timeline. Mice were divided into groups receiving either a 3 kHz fixed and dynamic up↑ tone or a 12 kHz fixed and dynamic down↓ tone. e. Tone frequency over time. f. Peri-stimulus time histogram (PSTH) showing average licking behaviors for the last 3 days of each condition (n = 9 mice). g. Average anticipatory lick rate (baseline subtracted) for 1 s preceding reward delivery (two-way ANOVA: long ITI vs short ITI F(1) = 9.3, **p = 0.0045). h. ANCCR simulation results from an 8 s dynamic cue followed by reward 1 s later for long ITI (teal) and short ITI (pink) conditions. Bold lines show the average of 20 iterations. i. Left, average dLight dopamine signals. Vertical dashed lines represent the ramp window from 3 to 8 s after cue onset, thereby excluding the influence of the cue onset and offset responses. Solid black lines show linear regression fit during window. Right, closeup of dopamine signal during window. j. Average peak dLight response to cue onset for LD and SD conditions (paired t-test: t(8) = 6.3, ***p = 2.3×10−4). k. dLight dopamine signal with linear regression fit during ramp window for example SD trials. Reported m is slope. l. Session average per-trial slope during ramp window for the first day and last 3 days of each condition (one-sided [last day LD < first day SD] paired t-test: t(8) = -2.1, *p = 0.036; one-sided [last day SD > first day SF] paired t-test: t(8) = 2.4, *p = 0.023). m. Average per-trial slope for last 3 days of each condition (Tukey HSD test: q = 3.8, LD vs SD **p = 0.0027, SD vs SF *p = 0.011). All data presented as mean ± SEM. See Supplementary Table 1 for full statistical details.

Given the speed with which dopamine ramps appeared and disappeared, we next tested whether the slope of dopamine ramps in the short ITI/dynamic tone condition depended on the previous ITI duration on a trial-by-trial basis. We found that there was indeed a statistically significant trial-by-trial correlation between the previous ITI duration and the current trial’s dopamine response slope in the short ITI/dynamic condition with ramps, but not in the long ITI/dynamic condition without ramps (Fig 2, Extended Data Fig 3). The dependence of a trial’s dopamine response slope with previous ITI was significantly negative, meaning that a longer ITI correlates with a weaker ramp on the next trial. This finding held when analyzing either animal-by-animal (Fig 2a-b) or the pooled trials across animals while accounting for mean animal-by-animal variability (Fig 2c). These results suggest that the eligibility trace time constant adapts rapidly to changing ITI in Pavlovian conditioning.

Per-trial dopamine ramps correlate with previous ITI.

a. Scatter plots showing the relationship between dopamine response slope within a trial and previous ITI for all trials in the last 3 days of SD condition. Each animal plotted individually with linear regression fit (black line). b. Linear regression β coefficients for previous ITI vs. trial slope calculated per animal (one-sided [< 0] one sample t-test: t(8) = -2.2, *p = 0.031). c. Scatter plot of Z-scored trial slope vs. previous ITI pooled across mice for all trials in the last 3 days of LD condition (linear regression: t(2671) = -4.1, R2 = 0.0063, ***p = 3.8 ×10−5). The Z-scoring per animal removes the effect of variable means across animals on the slope of the pooled data. All data presented as mean ± SEM. See Supplementary Table 1 for full statistical details.

We next tested whether the results from Pavlovian conditioning could be reproduced in an instrumental task. In keeping with prior demonstrations of dopamine ramps in head-fixed mice, we used a virtual reality (VR) navigational task in which head-fixed mice had to run towards a destination in a virtual hallway to obtain sucrose rewards (8, 10, 18, 42) (Fig 3a-b, Extended Data Fig 4). At reward delivery, the screen turned blank during the ITI and remained so until the next trial onset. After training animals in this task using a medium ITI, we changed the ITI duration to short or long for eight days before switching to the other (Fig 3c). We found evidence that mice learned the behavioral requirement during the trial period, as they significantly increased their running speed during trial onset (Fig 3d-e) and reached a similarly high speed prior to reward in both ITI conditions (Fig 3f-g). Consistent with the results from Pavlovian conditioning, the dopamine response to the onset of the hallway presentation was larger during the long ITI compared to the short ITI condition (Fig 3h-i), and dopamine ramps were observed only in the short ITI condition (Fig 3j-m). Unlike the Pavlovian conditioning, the change in the ITI resulted in a more gradual appearance or disappearance of ramps (Fig 3k), but there was still a weak overall correlation between dopamine response slope on a trial and the previous inter-reward interval (Extended Data Fig 5). These results are consistent with a more gradual change in the eligibility trace time constant in this instrumental task. Collectively, the core finding from Pavlovian conditioning that mesolimbic dopamine ramps are present only during short ITI conditions was reproduced in the instrumental VR task.

VR navigation dopamine ramps depend on ITI.

a. Head-fixed VR approach schematic. b. VR navigation task setup. Trials consisted of running down a patterned virtual hallway to receive sucrose reward. VR monitor remained black during the ITI. c. Experimental timeline. Following training, mice were assigned to either long or short ITI conditions for 8 days before switching. d. Velocity PSTH aligned to trial onset for long (teal) and short (pink) ITI conditions (n = 9 mice). e Average change in velocity at trial onset. Bottom asterisks indicate both conditions significantly differ from zero (one-sided [> 0] one sample t-test: long ITI t(8) = 6.4, ***p = 1.0 ×10−4 ; short ITI t(8) = 7.9, ***p = 2.3 ×10−5). Top asterisks indicate significant difference between conditions (paired t-test: t(8) = 4.3, **p = 0.0028). f. Velocity PSTH aligned to reward delivery. g. Average velocity during 1 s preceding reward (paired t-test: t(8) = 0.71, p = 0.50). h. PSTH showing average dLight dopamine signal aligned to trial onset. i. Comparison of peak dLight onset response (paired t-test: t(8) = 7.6, ***p = 6.3 ×10−5). j. Left, average dLight dopamine signal across distances spanning the entire virtual corridor. Vertical dashed lines represent the ramp window from 20 to 57 cm (10 cm before end of track). Solid black lines show linear regression fit during window. Right, closeup of dopamine signal during window. k. Session average per-trial slope during ramp window for all days of each condition. l. Scatter plot showing relationship between average per-session slope and inter-reward interval for the last 3 days in long (teal) and short (pink) ITI conditions. Black line indicates linear regression fit (linear regression: t(53) = -2.6, R2 = 0.12, *p = 0.012). m. Comparison of average per-trial slope during ramp window for last 3 days of both conditions (one-sided [long < short] paired t-test: t(8) = -2.1, *p = 0.035). All data presented as mean ± SEM. See Supplementary Table 1 for full statistical details.

Our results provide a general framework for understanding past results on dopamine ramps. According to ANCCR, the fundamental variable controlling the presence of ramps is the eligibility trace time constant. Based on first principles, this time constant depends on the ITI in common task designs (Supplementary Note 1). Thus, the ITI is a simple proxy to manipulate the eligibility trace time constant, thereby modifying dopamine ramps. In previous navigational tasks with dopamine ramps, there was no explicitly programmed ITI (4, 7, 13). As such, the controlling of the pace of trials by these highly motivated animals likely resulted in short effective ITI compared to trial duration. An instrumental lever pressing task with dopamine ramps similarly had no explicitly programmed ITI(9), and other tasks with observed ramps had short ITIs(8, 10, 12, 15, 16). One reported result that does not fit with a simple control of ramps by ITI is that navigational tasks produce weaker ramps with repeated training(7). These results are generally inconsistent with the stable ramps that we observed in Pavlovian conditioning across eight days (Fig 1l). A speculative explanation might be that when the timescales of events vary considerably (e.g., during early experience in instrumental tasks due to variability in action timing), animals use a short eligibility trace time constant to account for the potential non-stationarity of the environment. With repeated exposure, the experienced stationarity of the environment might increase the eligibility trace time constant, thereby complicating its relationship with the ITI. Alternatively, as suggested previously(7), repeated navigation may result in automated behavior that ignores the progress towards reward, thereby minimizing the calculation of associations of spatial locations with reward. Another set of observations superficially inconsistent with our assumption of the necessity of a sequence of external cues signaling proximity to reward for ramping to occur is that dopamine ramping dynamics can be observed even when only internal states signal reward proximity (e.g., timing a delayed action) (7, 12, 15). In these cases, however, animals were required to actively keep track of the passage of time, which therefore strengthens an internal progression of neural states signaling temporal proximity to reward. Once learned, these internal states could serve the role of external cues in the ANCCR framework (Supplementary Note 3).

Though the current experiments in this study were motivated by the ANCCR framework, they were not conducted to discriminate between theories. As such, though the data are largely consistent with ANCCR, they should not be treated as evidence explicitly for ANCCR (Supplementary Note 4). It may also be possible that these results can be explained by other theories of dopamine function. For instance, temporal discounting may depend on overall reward rate(23, 43, 44), and such changes in temporal discounting may affect predictions in alternative theories such as temporal difference value or RPE dopamine signaling. Regardless of such considerations, the current results provide a clear constraint for dopamine theories and demonstrate that an underappreciated experimental variable determines the emergence of mesolimbic dopamine ramps.

eLife Assessment

Floeder and colleagues report that dopamine ramps are determined by the duration of the intertrial interval of the task and the presence of dynamic cues that indicate reward proximity. The manuscript provides valuable new information on a topic of active discussion in the dopamine and reinforcement learning field and the strength of the evidence supporting the claims is solid.

Reviewer #1 (Public Review):

Summary:

In this study, Floedder et al report that dopamine ramps in both Pavlovian and Instrumental conditions are shaped by reward interval statistics. Dopamine ramps are an interesting phenomenon because at first glance they do not represent the classical reward prediction errors associated with dopamine signaling. Instead, they seem somewhat to bridge the gap between tonic and phasic dopamine, with an intense discussion still being held in the field about what is their actual behavioral role. Here, in tests with head-fixed mice, and dopamine being recorded with a genetically encoded fluorescent sensor in the nucleus accumbens, the authors find that dopamine ramps were only present when intertrial intervals were relatively short and the structure of the task (Pavlovian cue or progression in a VR corridor) contained elements that indicated progression towards the reward (e.g., a dynamic cue). The authors show that these findings are well explained by their previously published model of Adjusted Net Contingency of Causal Relation (ANCCR).

Strengths:

This descriptive study delineates some fundamental parameters that define dopamine ramps in the studied conditions. The short, objective, and to-the-point format of the manuscript is great and really does a service to potential readers. The authors are very careful with the scope of their conclusions, which is appreciated by this reviewer.

Weaknesses:

The discussion of the results is very limited to the conceptual framework of the authors' preferred model (which the authors do recognize, but it still is a limitation). The correlation analysis presented in panel I of Figure 3 seems unnecessary at best and could be misleading, as it is really driven by the categorical differences between the two conditions that were grouped for this analysis. There are some key aspects of the data and their relationship with each other, the previous literature, and the methods used to collect them, that could have been better discussed and explored.

Reviewer #2 (Public Review):

In this manuscript by Floeder et al., the authors report a correlation between ITI duration and the strength of a dopamine ramp occurring in the time between a predictive conditioned stimulus and a subsequent reward. They found this relationship occurring within two different tasks with mice, during both a Pavlovian task as well as an instrumental virtual visual navigation task. Additionally, they observed this relationship only in conditions when using a dynamic predictive stimulus. The authors relate this finding to their previously published model ANCCR in which the time constant of the eligibility trace is proportionate to the reward rate within the task.

The relationship between ITI duration and the extent of a dopamine ramp which the authors have reported is very intriguing and certainly provides an important constraint for models for dopamine function. As such, these findings are potentially highly impactful to the field. I do have a few questions for the authors which are written below.

(1) I was surprised to see a lack of counterbalance within the Pavlovian design for the order of the long vs short ITI. Ramping of the lick rate does increase from the long-duration ITIs to the short-duration ITI sessions. Although of course, this increase in ramping of the licking across the two conditions is not necessarily a function of learning, it doesn't lend support to the opposite possibility that the timing of the dynamic CS hasn't reached asymptotic learning by the end of the long-duration ITI. The authors do reference papers in which overtraining tends to result in a reduction of ramping, which would argue against this possibility, yet differential learning of the dynamic CS would presumably be required to observe this effect. Do the authors have any evidence that the effect is not due to heightened learning of the timing of the dynamic CS across the experiment?

(2) The dopamine response, as measured by dLight, seems to drop after the reward is delivered. This reduction in responding also tends to be observed with electrophysiological recordings of dopamine neurons. It seems possible that during the short ITI sessions, particularly on the shorter ITI duration trials, that dopamine levels may still be reduced from the previous trial at the onset of the CS on the subsequent trial. Perhaps the authors can observe the dynamics of the recovery of the dopamine response following a reward delivery on longer-duration ITIs in order to determine how quickly dopamine is recovering following a reward delivery. Are the trials with very short ITIs occurring within this period that dopamine is recovering from the previous trial? If so, how much of the effect may be due to this effect? It should be noted that the lack of observance of a ramp on the condition of short-duration ITIs with fixed CSs provides a potential control for this effect, yet the extent to which a natural ramp might occur following sucrose deliveries should be investigated.

(3) The authors primarily relate the finding of the correlation between the ITI and the slope of the ramp to their ANCCR model by suggesting that shorter time constants of the eligibility trace will result in more precisely timed predictors of reward across discrete periods of the dynamic cue. Based on this prediction, would the change in slope be more gradual, and perhaps be more correlated with a broader cumulative estimate of reward rate than just a single trial?

Reviewer #3 (Public Review):

Summary:

Floeder and colleagues measure dopamine signaling in the nucleus accumbens core using fiber photometry of the dLight sensor, in Pavlovian and instrumental tasks in mice. They test some predictions from a recently proposed model (ANCCR) regarding the existence of "ramps" in dopamine that have been seen in some previous research, the characteristics of which remain poorly understood.

They find that cues signaling a progression toward rewards (akin to a countdown) specifically promote ramping dopamine signaling in the nucleus accumbens core, but only when the intertrial interval just experienced was short. This work is discussed in the context of ongoing theoretical conceptions of dopamine's role in learning.

Strengths:

This work is the clearest demonstration to date of concrete training factors that seem to directly impact whether or not dopamine ramps occur. The existence of ramping signals has long been a feature of debates in the dopamine literature and this work adds important context to that. Further, as a practical assessment of the impact of a relatively simple trial structure manipulation on dopamine patterns, this work will be important for guiding future studies. These studies are well done and thoughtfully presented.

Weaknesses:

It remains somewhat unclear what limits are in place on the extent to which an eligibility trace is reflected in dopamine signals. In the current study, a specific set of ITIs was used, and one wonders if the relative comparison of ITI/history variables ("shorter" or "longer") is a factor in how the dopamine signal emerges, in addition to the explicit length ("short" or "long") of the ITI. Another experimental condition, where variable ITIs were intermingled, could perhaps help clarify some remaining questions.

In both tasks, cue onset responses are larger, and longer on long ITI trials. One concern is that this larger signal makes seeing a ramp during the cue-reward interval harder, especially with a fluorescence method like photometry. Examining the traces in Figure 1i - in the long, dynamic cue condition the dopamine trace has not returned to baseline at the time of the "ramp" window onset, but the short dynamic trace has. So one wonders if it's possible the overall return to baseline trend in the long dynamic conditions might wash out a ramp.

Not a weakness of this study, but the current results certainly make one ponder the potential function of cue-reward interval ramps in dopamine (assuming there is a determinable function). In the current data, licking behavior was similar on different trial types, and that is described as specifically not explaining ramp activity.

Author Response:

The following is the authors’ response to the original reviews.

Reviewer #1 (Public Review):

Summary:

In this study, Floedder et al report that dopamine ramps in both Pavlovian and Instrumental conditions are shaped by reward interval statistics. Dopamine ramps are an interesting phenomenon because at first glance they do not represent the classical reward prediction errors associated with dopamine signaling. Instead, they seem somewhat to bridge the gap between tonic and phasic dopamine, with an intense discussion still being held in the field about what is their actual behavioral role. Here, in tests with head-fixed mice, and dopamine being recorded with a genetically encoded fluorescent sensor in the nucleus accumbens, the authors find that dopamine ramps were only present when intertrial intervals were relatively short and the structure of the task (Pavlovian cue or progression in a VR corridor) contained elements that indicated progression towards the reward (e.g., a dynamic cue). The authors show that these findings are well explained by their previously published model of Adjusted Net Contingency of Causal Relation (ANCCR).

Strengths:

This descriptive study delineates some fundamental parameters that define dopamine ramps in the studied conditions. The short, objective, and to-the-point format of the manuscript is great and really does a service to potential readers. The authors are very careful with the scope of their conclusions, which is appreciated by this reviewer.

We thank the reviewer for their overall support of the formatting and scope of the manuscript.

Weaknesses:

The discussion of the results is very limited to the conceptual framework of the authors' preferred model (which the authors do recognize, but it still is a limitation). The correlation analysis presented in panel l of Figure 3 seems unnecessary at best and could be misleading, as it is really driven by the categorical differences between the two conditions that were grouped for this analysis. There are some key aspects of the data and their relationship with each other, the previous literature, and the methods used to collect them, that could have been better discussed and explored.

We agree with the reviewer that a weakness of the discussion was the limited framing of the results within the ANCCR model. To address this, we have expanded our introduction and discussion sections to provide a more thorough explanation of our model and possible leading alternatives.

We thank the reviewer for pointing out that Figure 3l may be misleading for readers; we removed this panel from the revised Figure 4.

We have further addressed the specific concerns raised by the reviewer in their comments to the authors. Indeed, we agree with the reviewer that the original manuscript was narrow in its focus regarding relationships between different aspects of the data. To more thoroughly explore how key variables – including dopamine ramp slope and onset response as well as licking behavior slope – could relate to each other, we have added Extended Data Figure 8. In this figure, we show that no correlations exist between any of these key variables in either dynamic tone condition; it is our hope that this additional analysis highlights the significance of the clear relationship between dopamine ramp slope and ITI duration.

Reviewer #2 (Public Review):

In this manuscript by Floeder et al., the authors report a correlation between ITI duration and the strength of a dopamine ramp occurring in the time between a predictive conditioned stimulus and a subsequent reward. They found this relationship occurring within two different tasks with mice, during both a Pavlovian task as well as an instrumental virtual visual navigation task. Additionally, they observed this relationship only in conditions when using a dynamic predictive stimulus. The authors relate this finding to their previously published model ANCCR in which the time constant of the eligibility trace is proportionate to the reward rate within the task.

The relationship between ITI duration and the extent of a dopamine ramp which the authors have reported is very intriguing and certainly provides an important constraint for models for dopamine function. As such, these findings are potentially highly impactful to the field. I do have a few questions for the authors which are written below.

We thank the reviewer for their interest in our findings and belief in their potential to be impactful in the field.

(1) I was surprised to see a lack of counterbalance within the Pavlovian design for the order of the long vs short ITI. Ramping of the lick rate does increase from the long-duration ITIs to the short-duration ITI sessions. Although of course, this increase in ramping of the licking across the two conditions is not necessarily a function of learning, it doesn't lend support to the opposite possibility that the timing of the dynamic CS hasn't reached asymptotic learning by the end of the long-duration ITI. The authors do reference papers in which overtraining tends to result in a reduction of ramping, which would argue against this possibility, yet differential learning of the dynamic CS would presumably be required to observe this effect. Do the authors have any evidence that the effect is not due to heightened learning of the timing of the dynamic CS across the experiment?

We appreciate the reviewer expressing their surprise regarding the lack of counterbalance in our Pavlovian experimental design. We previously did not explicitly do this because the ramps disappeared in the short ITI/fixed tone condition, indicating that their presence is not just a matter of total experience in the task. However, we agree that this is incidental, but not direct evidence. To address this drawback, we repeated the Pavlovian experiment in a new cohort of animals with a revised training order, switching conditions such that the short ITI/dynamic tone (SD) condition preceded the long ITI/dynamic tone (LD) condition (see revised Figure 2a). Despite this change in the training order, the main findings remain consistent: positive dLight slopes (i.e., dopamine ramps) are only observed in the SD condition (Figure 2b-d).

We thank the reviewer for raising these questions regarding licking behavior and learning and their relationship with dopamine ramps. Indeed, a closer look at the average licking behavior reveals subtle differences across conditions (Figure 1f and Extended Data Figure 5a). While the average lick rate during the ramp window does not differ across conditions (Extended Data Figure 5c), the ramping of the lick rate during this window is higher for dynamic tone conditions compared to fixed tone conditions (Extended Data Figure 5d). Despite these differences, we still believe that the main comparison between the dopamine slope in the SD vs LD condition remains valid given their similar lick ramping slopes. Furthermore, our primary measure of learning is not lick slope, but anticipatory lick rate during the 1 s trace preceding reward delivery, which is robustly nonzero across cohorts and conditions (Figure 1g and Extended Data Figure 5b).

Taken together, we hope that the results from our counterbalanced Pavlovian training and more rigorous analysis of lick behavior across conditions provide sufficient evidence to assuage concerns that the differences in ramping dopamine simply reflect differences in learning.

(2) The dopamine response, as measured by dLight, seems to drop after the reward is delivered. This reduction in responding also tends to be observed with electrophysiological recordings of dopamine neurons. It seems possible that during the short ITI sessions, particularly on the shorter ITI duration trials, that dopamine levels may still be reduced from the previous trial at the onset of the CS on the subsequent trial. Perhaps the authors can observe the dynamics of the recovery of the dopamine response following a reward delivery on longer-duration ITIs in order to determine how quickly dopamine is recovering following a reward delivery. Are the trials with very short ITIs occurring within this period that dopamine is recovering from the previous trial? If so, how much of the effect may be due to this effect? It should be noted that the lack of observance of a ramp on the condition of shortduration ITIs with fixed CSs provides a potential control for this effect, yet the extent to which a natural ramp might occur following sucrose deliveries should be investigated.

We thank the reviewer for highlighting the possibility that ramps may be due to the dopamine response recovery following reward delivery. Given that peak reward dopamine responses tend to be larger in long ITI conditions, however, we felt that it was inappropriate to compare post-reward dopamine recovery times across conditions. Instead, we decided to directly compare the dLight slope 2s before cue onset (“pre-cue window,” a proxy for recovery from previous trial) with the dLight slope during our ramp window from 3 to 8s after cue onset (Extended Data Figure 6a). There were no significant differences in pre-cue dLight slope across conditions (Extended Data Figure 6b); this suggests that the ramping slopes seen in the SD condition, but not other conditions, is not simply due to the natural dopamine recovery response following reward delivery. Furthermore, if the dopamine ramps observed in the SD condition were a continuation of the post-reward dopamine recovery from the previous trial, we would expect to see a positive correlation between the dLight slope before and during the cue. However, there is no such correlation between the dLight slopes in the ramp window vs. pre-cue window in the SD condition (Extended Data Figure 6c-d). We believe that this observation, along with the builtin control of the SF condition mentioned by the reviewer, serves as evidence against the possibility of our ramp results being due to a natural ramp after reward delivery.

(3) The authors primarily relate the finding of the correlation between the ITI and the slope of the ramp to their ANCCR model by suggesting that shorter time constants of the eligibility trace will result in more precisely timed predictors of reward across discrete periods of the dynamic cue. Based on this prediction, would the change in slope be more gradual, and perhaps be more correlated with a broader cumulative estimate of reward rate than just a single trial?

To clarify, we do not propose that a smaller eligibility trace time constant results in more precise timing per se. Instead, we believe that the rapid eligibility trace decay from smaller time constants gives greater causal predictive power for later periods in the dynamic cue (see Extended Data Figure 1) since the memory of the earlier periods of the cue is weaker.

We appreciate the reviewer’s curiosity regarding the influence of a broader cumulative estimate of reward vs. only the immediately preceding ITI on dopamine ramp slopes. Indeed, in several instrumental tasks (e.g., Krausz et al., Neuron, 2023), recent reward rate modulates the magnitude of dopamine ramps, making this an important variable to investigate. We chose to use linear regression for each mouse separately to analyze the relationship between the trial dopamine slope and the average previous ITI for the past 1 through 10 most recent trials. In the SD condition, as reported in our earlier manuscript, there was a significantly negative dependence of trial dopamine slope with the single previous ITI (i.e., if the previous ITI was long, the next trial tends to have a weaker ramp). This negative dependence, however, only held for a single previous trial; there was no clear relationship between the per-trial dopamine slope and the average of the past 2 through 10 ITIs (Extended Data Figure 7a). For the LD condition, on the other hand, there is no clear relationship between the per-trial dopamine slope and the average previous ITI for any of the past 1 through 10 trials, with one exception: there is a significantly negative dependence of trial dopamine slope with the average ITI of the previous 2 trials (Extended Data Figure 7b). This longer timescale relationship in the LD condition suggests that the adaptation of the eligibility trace time constant is nuanced and depends on the general ITI length.

In general, though we reason that the eligibility trace time constant should depend on overall event rates, we do not currently propose a real-time update rule for the eligibility trace time constant depending on recent event rates. Accordingly, we are currently agnostic about the actual time scale of history of recent event rate calculation that mediates the eligibility trace time constant. Our experimental results suggest that when the ITI is generally short for Pavlovian conditioning, the eligibility trace time constant adapts to ITI on a rapid timescale. However, only a small fraction of the variability of this rapid fluctuation is captured by recent ITI history. A more thorough investigation of this real-time update rule would need to be done in the future.

Reviewer #3 (Public Review):

Summary:

Floeder and colleagues measure dopamine signaling in the nucleus accumbens core using fiber photometry of the dLight sensor, in Pavlovian and instrumental tasks in mice. They test some predictions from a recently proposed model (ANCCR) regarding the existence of "ramps" in dopamine that have been seen in some previous research, the characteristics of which remain poorly understood.

They find that cues signaling a progression toward rewards (akin to a countdown) specifically promote ramping dopamine signaling in the nucleus accumbens core, but only when the intertrial interval just experienced was short. This work is discussed in the context of ongoing theoretical conceptions of dopamine's role in learning.

Strengths:

This work is the clearest demonstration to date of concrete training factors that seem to directly impact whether or not dopamine ramps occur. The existence of ramping signals has long been a feature of debates in the dopamine literature and this work adds important context to that. Further, as a practical assessment of the impact of a relatively simple trial structure manipulation on dopamine patterns, this work will be important for guiding future studies. These studies are well done and thoughtfully presented.

We thank the reviewer for recognizing the context that our study adds to the dopamine literature and the potential for our experiments to guide future work.

Weaknesses:

It remains somewhat unclear what limits are in place on the extent to which an eligibility trace is reflected in dopamine signals. In the current study, a specific set of ITIs was used, and one wonders if the relative comparison of ITI/history variables ("shorter" or "longer") is a factor in how the dopamine signal emerges, in addition to the explicit length ("short" or "long") of the ITI. Another experimental condition, where variable ITIs were intermingled, could perhaps help clarify some remaining questions.

Though we used ITIs of fixed means, due to the exponential nature of their distribution, we did intermingle ITIs of various durations in both our long and short ITI conditions. The distribution of ITI durations is visualized in Figure 1c for Pavlovian conditioning and Extended Data Figure 9b for VR navigation.

The relative comparison between consecutive ITIs was not something we originally explored, so we thank the reviewer for wondering how it impacts the dopamine signal. To investigate this, we quantified both the change in ITI (+ or - Δ ITI for relatively longer or shorter, respectively) and the change in dopamine ramp slope between consecutive trials in the SD condition (Figure 3d). Across each mouse separately, we found a significantly negative relationship between Δ slope and Δ ITI (Figure 3e-f). Also, the average Δ slope was significantly greater for consecutive trials with a Δ ITI below -1 s compared to trials with a Δ ITI above +1 s (Figure 3g). Altogether, these findings suggest that relative comparison of ITIs does correlate with changes in the dopamine signal; a relatively longer ITI tends to have a weaker ramp, which fits in nicely with the expected inverse relationship between ITI and dopamine ramp slope from our ANCCR model.

In both tasks, cue onset responses are larger, and longer on long ITI trials. One concern is that this larger signal makes seeing a ramp during the cue-reward interval harder, especially with a fluorescence method like photometry. Examining the traces in Figure 1i - in the long, dynamic cue condition the dopamine trace has not returned to baseline at the time of the "ramp" window onset, but the short dynamic trace has. So one wonders if it's possible the overall return to baseline trend in the long dynamic conditions might wash out a ramp.

This is a good point, and we thank the reviewer for raising it. Certainly, the cue onset response is significantly larger in long ITI conditions (see Figure 1i-j and Figure 4h-j). To avoid any bleed over effect, we intentionally chose ramp window periods during later portions of the trial (in line with work from others e.g., Kim et al., Cell, 2020). While the cue onset dopamine pulse seems to have flatlined by the start of the ramp window period, the dopamine levels clearly remain elevated relative to pre-cue baseline. This type of signal has been observed with fiber photometry in other Pavlovian conditioning paradigms with long cue durations (e.g., Jeong et al., Science, 2022). Because of the persistently elevated dopamine levels, it is certainly possible that a ramping signal during the cue is getting washed out; with the bulk fluorescence photometry technique we employed in this study, this possibility is unfortunately difficult to completely rule out. However, the long ITI/fixed tone (LF) condition could serve as a potential control given the overall similarity in the dopamine signal between the LF and LD conditions: both conditions have large cue onset responses with elevated dopamine throughout the duration of the cue (see Extended Data Figures 2c and 3c). Critically, the LD condition lacks a noticeable ramp despite the dynamic tone providing information on temporal proximity to reward, which is thought to be necessary for dopamine ramps to occur. Importantly, regardless of whether a ramp is masked in the long ITI dynamic condition, most studies investigate such a condition in isolation and would report the absence of dopamine ramps. Thus, at a descriptive level, we believe it remains true that observable dopamine ramps are only present when the ITI is short.

Not a weakness of this study, but the current results certainly make one ponder the potential function of cue-reward interval ramps in dopamine (assuming there is a determinable function). In the current data, licking behavior was similar on different trial types, and that is described as specifically not explaining ramp activity.

We agree that this work naturally raises the question of the function of dopamine ramps. However, selective and precise manipulation of only the dopamine ramps without altering other features such as phasic responses, or inducing dopamine dips, is highly technically challenging at this moment; due to this challenge, we intentionally focused on the conditions that determine the presence or absence of dopamine ramps rather than their function. We agree with the reviewer that studying the specific function of dopamine ramps is an interesting future question.

Reviewing Editor:

The reviewers felt the results are of considerable and broad interest to the neuroscience community, but that the framing in terms of ANCCR undermined the scope of the findings as did the brief nature of the formatting of the manuscript. In addition, the reviewers felt that the relationship between ramp dynamics, behavior, and ITI conditions requires more in-depth analyses. Relatedly, the lack of counterbalancing of the ITI durations was considered to be a drawback and needs to be addressed as it may affect the baseline. Addressing these issues in a satisfactory manner would improve the assessment of the manuscript to important/convincing.

We truly appreciate the valuable feedback provided on this manuscript by all three reviewers and the reviewing editor. Based on this input, we have significantly revised the manuscript to address the issues brought up by the reviewers. Firstly, we have conducted additional experiments to counterbalance the ITI conditions for Pavlovian conditioning; this strengthened our results by confirming our original findings that ITI duration, rather than training order, is the key variable controlling the presence or absence of dopamine ramps. Secondly, we completed more rigorous analyses to further explore the relationship between dopamine dynamics, animal behavior, and ITI duration; we generally found no significant correlations between these variables, with a notable exception being our main finding between ITI duration and dopamine ramp slope. Finally, we revised and expanded our writing to both explain predictions from our ANCCR model in less technical language and explore how alternative theoretical frameworks could potentially explain our findings. In doing so, we hope that our manuscript is now more accessible and of interest to a broad audience of neuroscience readers.

Reviewer #1 (Recommendations For The Authors):

The study could be improved if the authors performed a more detailed comparison of how other theoretical frameworks, beyond ANCCR could account for the observed findings. Also, the correlation analysis presented in the panel I of Figure 3 seems unnecessary and potentially spurious, as the slope of the correlation is clearly mostly driven by the categorical differences between the two ITI conditions, which were combined for the analysis - it's not clear what is the value of this analysis beyond the group comparison presented in the following panel.

Again, we thank the reviewer for elaborating on their concern regarding Figure 3l – we have removed it from the revised Figure 4.

The relationship between ramp dynamics with the behavior and the large differences in cue onset responses between short and long ITI conditions could have been better explored. If I understand correctly the overarching proposal of this and other publications by this group, then the differences in cue responses is determined by the spacing of rewards in a somewhat similar way that the ramps are. So, is there a trial-by-trial correlation between the amplitude of the cue responses and the slope of the ramps? Is there a correlation between any of these two measures with the licking behavior, and if so, does it change with the ITI condition? A more thorough exploration of these relationships would help support the proposal of the primacy of inter-event spacing in determining the different types of dopamine responses in learning.

There are certainly interesting relationships between dopamine dynamics, behavior, and ITI that we failed to explore in our original manuscript – we appreciate the reviewer bringing them up. We found no correlation between dopamine ramp slope and cue onset response in either the SD or LD condition (Extended Data Fig 8a-b). Moreover, we found no correlation between either of these variables and the trial-by-trial licking behavior (Extended Data Fig 8c-f). Finally, there is no relationship between licking behavior and previous ITI duration (Extended Data Fig 8g-h), suggesting that behavioral differences do not account for differences in the dopamine ramp slope. Together, the lack of significant relationships between these other variables highlights the specific, clear relationship between ITI duration and dopamine ramp slope.

Finally, another issue I feel could have been better discussed is how the particular settings of both tasks might be biasing the results. For example, there is an issue to be considered about how the dopamine ramp dynamics reported here, especially the requirement of a dynamic cue for ramps to be present, square with the previous published results by one of the authors - Mohebi et al, Nature, 2019. In that manuscript, rats were executing a bandit task where, to this reviewer's understanding, there was no explicit dynamic cue aside from the standard sensory feedback of the rats moving around in the behavior boxes to approach a nose poke port. Is the idea that this sensory feedback could function as a dynamic cue? If that's the case, then this short-scale, movement-related feedback should also function as a dynamic cue in a freely moving Pavlovian condition, when the animals must also move towards a reward delivery port, right? Therefore, could it be that the experimental "requirement" of a dynamic cue is only present in a head-fixed condition? One could phrase this in a different way to Steelman and potentially further the authors' proposal: perhaps in any slightly more naturalistic setting, the interaction of the animals with their environment always functions as a dynamic cue indicating proximity to reward, and this relationship was experimentally isolated by the use of head fixation (but not explicitly compared with a freely moving condition) in the present study. I think that would be an interesting alternative to consider and discuss, and perhaps explore experimentally at some point.

We thank the reviewer for raising this important point regarding the influence of our experimental settings on our results. At first glance, it could appear that our results demonstrating the necessity of a dynamic cue for ramps in a head-fixed setting do not fit neatly with other results in a freely moving setup (e.g., Collins et al., Scientific Reports, 2016; Mohebi et al., Nature, 2019). Exactly as the reviewer states though, we believe that sensory feedback from the environment in freely moving preparations serves the same function as a dynamic progression of cues. We have considered the implications of methodological differences between head-fixed and freely moving preparations in the discussion section.

Reviewer #2 (Recommendations For The Authors):

This comment relates indirectly to comment 3, in that the authors intermix theory throughout the manuscript. I think this would be fine if the experiment was framed directly in terms of ANCCR, but the authors specifically mention that this experiment wasn't developed to distinguish between different theories. As such, it seems difficult to assess the scope of the comments regarding theory within the paper because they tend to be specifically related to ANCCR. For instance, the last comment has broad implications of how the ramp might be related to the overall reward rate, an interesting finding that constrains classes of dopamine models rather than evidence just for ANCCR. Perhaps adding a discussion section that allows the authors to focus more on theory would be beneficial for this manuscript.

We appreciate this suggestion by the reviewer. We have updated both our introduction and discussion sections to elaborate more thoroughly on theory.

Reviewer #3 (Recommendations For The Authors):

The paper could potentially benefit from the use of more accessible language to describe the conceptual basis of the work, and the predictions, and a bit of reformatting away from the brief structure with lots of supplemental discussion.

For example, in the introduction, the line - "Varying the ITI was critical because our theory predicts that the ITI is a variable controlling the eligibility trace time constant, such that a short ITI would produce a small time constant relative to the cue-reward interval (Supplementary Note 1)". As far as I can tell, this is meant to get across the notion that dopamine represents some aspect of the time between rewards - dopamine signals will differ for cues following short vs long intervals between rewards.

As written, the language of the paper takes a fair bit of parsing, but the notions are actually pretty simple. This is partly due to the brief format the paper is written in, where familiarity with the previous papers describing ANCCR is assumed.

From a readability standpoint, and the potential impact of the paper on a broad audience, perhaps this could be considered as a point for revision.

We thank the reviewer for pointing out the drawbacks of our technical language and brief formatting. To address this, we have removed the majority of the supplementary notes and expanded our introduction and discussion sections. In doing so, we hope that the conceptual foundations of this work, and potential alternative theoretical explanations, are accessible and impactful for a broad audience of readers.

Version 2

Abstract

Abstract

Mesolimbic dopamine activity occasionally exhibits ramping dynamics, reigniting debate on theories of dopamine signaling. This debate is ongoing partly because the experimental conditions under which dopamine ramps emerge remain poorly understood. Here, we show that during Pavlovian and instrumental conditioning, mesolimbic dopamine ramps are only observed when the inter-trial interval is short relative to the trial period. These results constrain theories of dopamine signaling and identify a critical variable determining the emergence of dopamine ramps.

Introduction

Mesolimbic dopamine activity was classically thought to operate in either a “phasic” or a “tonic” mode(1–3). Yet, recent evidence points to a “quasi-phasic” mode in which mesolimbic dopamine activity exhibits ramping dynamics(4–16). This discovery reignited debate on theories of dopamine function because it appeared inconsistent with the dominant theory that dopamine signaling conveys temporal difference reward prediction error (RPE)(17), since ramping dopamine would paradoxically be a “predictable prediction error”(18). Recent work has hypothesized that dopamine ramps reflect the value of ongoing states, serving as a motivational signal(1, 4–6). Others have argued that ramping dopamine indeed reflects RPE under some assumptions, namely correction of uncertainty via sensory feedback(11, 15, 19), representational error(20), or memory lapses(21). Still others have proposed that dopamine ramps reflect a causal influence of actions on rewards in instrumental tasks(12). This debate has been exacerbated in part because there is no clear understanding of why dopamine ramps appear only under some experimental conditions. Accordingly, uncovering a unifying principle of the conditions under which dopamine ramps appear will provide important constraints on theories of dopamine function(1, 12, 18–42).

To investigate the necessary conditions for dopamine ps, we turned to our recent work proposing that dopamine acts as a teaching signal for causal learning by representing the Adjusted Net Contingency for Causal Relations (ANCCR)(32, 42, 43). The crux of the ANCCR model is that learning to predict a future meaningful event (e.g., reward) can occur by looking back in time for potential causes of that event. Critically, the ability to learn by looking backwards depends on how long one holds on to past events in memory. If the past is quickly forgotten, there is little ability to identify causes that occurred long before a reward. On the other hand, maintaining memory for too long is computationally inefficient and would allow illusory associations across long delays. Thus, for optimal learning, the timescale for memory maintenance should flexibly depend on “environmental timescales” set by the overall rates of events. In ANCCR, this is achieved by controlling the duration of a memory trace of past events with the “eligibility trace” time constant (illustrated in Extended Data Fig 1). When there is a high average rate of events, the eligibility trace time constant is small. Accordingly, we successfully simulated dopamine ramping dynamics assuming two conditions: a dynamic progression of cues that signal temporal proximity to reward, and a small eligibility trace time constant relative to the trial period(32). However, whether these conditions are sufficient to experimentally produce mesolimbic dopamine ramps in vivo remains untested.

In this study, we designed experiments to address the influence of environmental timescales on dopamine ramps. Specifically, we sought to test the key prediction that dopamine ramps would be observed for a small, but not large, eligibility trace time constant, which is hypothesized to emerge for high overall event rate. To do so, we manipulated the intertrial interval (ITI) duration in both an auditory Pavlovian conditioning paradigm and a virtual reality navigation task. The results confirmed our key prediction from ANCCR, providing a clear constraint on theoretical explanations for the controversial phenomenon of dopamine ramps.

Results

We first measured mesolimbic dopamine release in the nucleus accumbens core using a dopamine sensor (dLight1.3b)(44) in an auditory cue-reward task. We varied both the presence or absence of a progression of cues indicating reward proximity (“dynamic” vs “fixed” tone) and the inter-trial interval (ITI) duration (short vs long ITI). Varying the ITI was critical because our theory predicts that the ITI is a variable controlling the eligibility trace time constant, such that a short ITI would produce a small time constant relative to the cue-reward interval (Supplementary Note 1, Fig 1a-e). In all four experimental conditions, head-fixed mice learned to anticipate the sucrose reward, as reflected by anticipatory licking (Fig 1f-g). In line with our earlier work, we showed that simulations of ANCCR exhibit a larger cue onset response when the ITI is long and exhibit ramps only when the ITI is short (Fig 1h). Consistent with these simulated predictions, experimentally measured mesolimbic dopamine release had a much higher cue onset response for long ITI (Fig 1i-j). Furthermore, dopamine ramps were observed only when the ITI was short and the tone was dynamic (Fig 1i, k-m, Extended Data Fig 2). Indeed, dopamine ramps—quantified by a positive slope of dopamine response vs time within trial over the last five seconds of the cue—appeared on the first day after transition from a long ITI/dynamic tone condition to a short ITI/dynamic tone condition and disappeared on the first day after transition from a short ITI/dynamic tone condition to a short ITI/fixed tone condition (Fig 1l). These results confirm the key prediction of our theory in Pavlovian conditioning.

Pavlovian conditioning dopamine ramps depend on ITI.

a. Top, fiber photometry approach schematic for nucleus accumbens core (NAcC) dLight recordings. Bottom, head-fixed mouse. b. Pavlovian conditioning experimental setup. Trials consisted of an 8 s auditory cue followed by sucrose reward delivery 1 s later. c. Cumulative Distribution Function (CDF) of ITI duration for long (solid line, mean 55 s) and short ITI (dashed line, mean 8 s) conditions. d. Experimental timeline. Mice were divided into groups receiving either a 3 kHz fixed and dynamic up↑ tone or a 12 kHz fixed and dynamic down↓ tone. e. Tone frequency over time. f. Peri-stimulus time histogram (PSTH) showing average licking behaviors for the last 3 days of each condition (n = 9 mice). g. Average anticipatory lick rate (baseline subtracted) for 1 s preceding reward delivery (two-way ANOVA: long ITI vs short ITI F(1) = 9.3, **p = 0.0045). h. ANCCR simulation results from an 8 s dynamic cue followed by reward 1 s later for long ITI (teal) and short ITI (pink) conditions. Bold lines show the average of 20 iterations. i. Left, average dLight dopamine signals. Vertical dashed lines represent the ramp window from 3 to 8 s after cue onset, thereby excluding the influence of the cue onset and offset responses. Solid black lines show linear regression fit during window. Right, closeup of dopamine signal during window. j. Average peak dLight response to cue onset for LD and SD conditions (paired t-test: t(8) = 6.3, ***p = 2.3×10−4). k. dLight dopamine signal with linear regression fit during ramp window for example SD trials. Reported m is slope. l. Session average per-trial slope during ramp window for the first day and last 3 days of each condition (one-sided [last day LD < first day SD] paired t-test: t(8) = −2.1, *p = 0.036; one-sided [last day SD > first day SF] paired t-test: t(8) = 2.4, *p = 0.023). m. Average per-trial slope for last 3 days of each condition (Tukey HSD test: q = 3.8, LD vs SD **p = 0.0027, SD vs SF *p = 0.011). All data presented as mean ± SEM. See Supplementary Table 1 for full statistical details.

While these results are consistent with the idea that dopamine ramps are shaped by the ITI, an alternative explanation could be differences in behavioral learning across experimental conditions. To test this possibility, we repeated the same Pavlovian conditioning paradigm with a counterbalanced training order in a second cohort of mice (Fig 2a). Despite the shuffled training order, this cohort behaved similarly and showed robust anticipatory licking across conditions (Extended Data Fig 3a-b). As with the previous cohort, dopamine ramps were only observed in the short ITI/dynamic tone condition, rapidly appearing on the first day of this condition and disappearing on the first day of the subsequent long ITI/dynamic tone condition (Fig 2b-d, Extended Data Fig 4, Extended Data Fig 5). Critically, the presence of dopamine ramps during the last five seconds of the cue could not be explained by variations in behavior; during this period, anticipatory licking was similar across all conditions, and there was no difference in the slope of the lick rate between the dynamic tone conditions (Extended Data Fig 3c-d). Taken together, these results rule out any effects of differential learning across conditions on dopamine ramps.

Pavlovian conditioning dopamine ramps do not depend on training order.

a. Experimental timeline in which the SD condition occurs before the LD condition. b. Left, average dLight dopamine signals for SD and LD conditions. Vertical dashed lines represent the ramp window from 3 to 8 s after cue onset. Solid black lines show linear regression fit during window. Right, closeup of dopamine signal during window (n = 9 mice). c.Session average per-trial slope during ramp window for the first day and last 3 days of each condition (last day LF to first day SD: **p = 0.0046, last day SD to first day LD: **p = 0.0067). d. Average per-trial slope for last 3 days of each condition (LF vs SD: ***p = 9.1 ×10−4, SD vs LD: *p = 0.010).

Although the difference in dopamine ramp slope seems to be well explained by the ITI condition, it might instead reflect differences in post-reward dopamine dynamics, which drop below baseline. As dopamine levels recover from this drop over several seconds, it could appear as a dopamine ramp on the subsequent trial given a sufficiently short ITI. The lack of dopamine ramps in the short ITI/fixed tone condition serves as a control for this, however (Fig 1l-m). Furthermore, there is no significant difference in the pre-cue dopamine slope between conditions nor is there a correlation between the precue dopamine slopes and the dopamine ramp slopes during the cue in the short ITI/dynamic condition (Extended Data Fig 6). As such, our results cannot be captured by a natural ramp in the dopamine signal following reward.

Given the speed with which dopamine ramps appeared and disappeared, we next tested whether the slope of dopamine ramps in the short ITI/dynamic tone condition depended on the previous ITI duration on a trial-by-trial basis. We found that there was indeed a statistically significant trial-by-trial correlation between the previous ITI duration and the current trial’s dopamine response slope in the short ITI/dynamic condition with ramps, but not in the long ITI/dynamic condition without ramps (Fig 3a-c). The dependence of a trial’s dopamine response slope with previous ITI was significantly negative, meaning that a longer ITI correlates with a weaker ramp on the next trial. This finding held when analyzing either animal-by-animal (Fig 3a-b) or the pooled trials across animals while accounting for mean animal-by-animal variability (Fig 3c). This relationship was only significant for a single previous trial, however, and did not hold for a broader estimate of average previous ITIs (Extended Data Fig 7). In addition, we quantified how the relative change in ITI duration between consecutive trials correlates with changes in dopamine ramp slope (Fig 3d). We found a significantly negative relationship between the change in dopamine slope and change in ITI (Fig 3e-f). Furthermore, the change in slope was significantly greater for relative decreases in ITI compared to relative increases in ITI, indicating that a relatively shorter ITI tends to have a stronger ramp (Fig 3g). These results suggest that the eligibility trace time constant adapts rapidly to changing ITI in Pavlovian conditioning.

Per-trial dopamine ramps correlate with previous ITI.

a. Scatter plot for an example animal showing the relationship between dopamine response slope within a trial and previous ITI for all trials in the last 3 days of SD condition. Plotted with linear regression fit (black line) used to find this animal’s β coefficient of −0.045. b. Linear regression β coefficients for previous ITI vs. trial slope calculated per animal (***p = 5.6 ×10−4). c. Scatter plot of Z-scored trial slope vs. previous ITI pooled across mice for all trials in the last 3 days of SD condition (***p = 6.6 ×10−10). The Z-scoring per animal removes the effect of variable means across animals on the slope of the pooled data. d.dLight dopamine signal for two consecutive example SD trials showing the change in ITI and change in slope. The grey shaded regions indicate ITIs, and the vertical dashed lines mark the ramp window period. Reported m is slope. e. Scatter plot for the same example animal in a showing the relationship between the change in dopamine slopes and the chance in ITI across all trials in the last 3 days of SD condition. Plotted with linear regression fit (black line). Dot colors indicate magnitude of Δ ITI: light pink for Δ ITI below −1 s; grey for Δ ITI between −1 s and 1 s; dark pink for Δ ITI above 1 s. f. Linear regression β coefficients for Δ ITI vs. Δ slope calculated per animal (***p = 3.2 ×10−4). g. Comparison of the average Δ slope for Δ ITI below −1 s vs above 1 s (*** p = 2.3 ×10−4).

Due to the robust relationship between ITI and dopamine ramp slope on a per-trial basis, we next sought to explore the potential relationships between other important dopaminergic and behavioral variables. Though the dopamine cue onset response is significantly greater in the long compared to short ITI/dynamic tone condition, there is no apparent relationship between the cue onset response and dopamine ramp slope in either condition (Extended Data Fig 8a-b). Furthermore, neither the cue onset response nor the dopamine ramp slope correlates with the per-trial behavior quantified as lick slope (Extended Data Fig 8c-f). Finally, unlike the ramping dopamine slope, this ramping lick slope did not correlate with ITI duration (Extended Data Fig 8g-h). The fact that this exploration of additional variables yielded no significant relationships highlights the unique, specific influence of ITI on dopamine ramp slope.

We next tested whether the results from Pavlovian conditioning could be reproduced in an instrumental task. In keeping with prior demonstrations of dopamine ramps in head-fixed mice, we used a virtual reality (VR) navigational task in which head-fixed mice had to run towards a destination in a virtual hallway to obtain sucrose rewards(11, 15, 19, 45) (Fig 4a-b, Extended Data Fig 9). At reward delivery, the screen turned blank during the ITI and remained so until the next trial onset. After training animals in this task using a medium ITI, we changed the ITI duration to short or long for eight days before switching to the other (Fig 4c). We found evidence that mice learned the behavioral requirement during the trial period, as they significantly increased their running speed during trial onset (Fig 4d-e) and reached a similarly high speed prior to reward in both ITI conditions (Fig 4f-g). Consistent with the results from Pavlovian conditioning, the dopamine response to the onset of the hallway presentation was larger during the long ITI compared to the short ITI condition (Fig 4h-i), and dopamine ramps were observed only in the short ITI condition (Fig 4j-m). Unlike the Pavlovian conditioning, the change in the ITI resulted in a more gradual appearance or disappearance of ramps (Fig 4l), but there was still a weak overall correlation between dopamine response slope on a trial and the previous inter-reward interval (Extended Data Fig 10). These results are consistent with a more gradual change in the eligibility trace time constant in this instrumental task. Collectively, the core finding from Pavlovian conditioning that mesolimbic dopamine ramps are present only during short ITI conditions was reproduced in the instrumental VR task.

VR navigation dopamine ramps depend on ITI.

a. Head-fixed VR approach schematic. b. VR navigation task experimental setup. Trials consisted of running down a patterned virtual hallway to receive sucrose reward. VR monitor remained black during the ITI. c. Experimental timeline. Following training, mice were assigned to either long or short ITI conditions for 8 days before switching. d. Velocity PSTH aligned to trial onset for long (teal) and short (pink) ITI conditions (n = 9 mice). e Average change in velocity at trial onset. Bottom asterisks indicate both conditions significantly differ from zero (long: ***p = 1.0 ×10−4; short: ***p = 2.3 ×10−5). Top asterisks indicate significant difference between conditions (**p = 0.0028). f. Velocity PSTH aligned to reward delivery. g. Average velocity during 1 s preceding reward (p = 0.50). h. PSTH showing average dLight dopamine signal aligned to trial onset. i. Comparison of peak dLight onset response (***p = 6.3 ×10−5). j. Left, average dLight dopamine signal across distances spanning the entire virtual corridor. Vertical dashed lines represent the ramp window from 20 to 57 cm (10 cm before end of track). Solid black lines show linear regression fit during window. Right, closeup of dopamine signal during window. k. dLight dopamine signal with linear regression fit during ramp window for example short ITI trials. Reported m is slope. l. Session average per-trial slope during ramp window for all days of each condition. m. Comparison of average per-trial slope during ramp window for last 3 days of both conditions (*p = 0.035).

Discussion

Our results provide a general framework for understanding past results on dopamine ramps. According to ANCCR, the fundamental variable controlling the presence of ramps is the eligibility trace time constant. Based on first principles, this time constant depends on the ITI in common task designs (Supplementary Note 1). Thus, the ITI is a simple proxy to manipulate the eligibility trace time constant, thereby modifying dopamine ramps. In previous navigational tasks with dopamine ramps, there was no explicitly programmed ITI(4, 10, 16). As such, the controlling of the pace of trials by these highly motivated animals likely resulted in short effective ITI compared to trial duration. An instrumental lever pressing task with dopamine ramps similarly had no explicitly programmed ITI(7), and other tasks with observed ramps had short ITIs(8, 11, 13–15). One reported result that does not fit with a simple control of ramps by ITI is that navigational tasks produce weaker ramps with repeated training(10). These results are generally inconsistent with the stable ramps that we observed in Pavlovian conditioning across eight days (Fig 1l). A speculative explanation might be that when the timescales of events vary considerably (e.g., during early experience in instrumental tasks due to variability in action timing), animals use a short eligibility trace time constant to account for the potential non-stationarity of the environment. With repeated exposure, the experienced stationarity of the environment might increase the eligibility trace time constant, thereby complicating its relationship with the ITI. Alternatively, as suggested previously(7, 10), repeated navigation may result in automated behavior that ignores the progress towards reward, thereby minimizing the calculation of associations of spatial locations with reward.

While the focus of this study was environmental timescales set by the ITI, we also assumed that a dynamic sequence of external cues signaling temporal or spatial proximity to reward would be required for dopamine ramps to occur. Indeed, our results that dopamine ramps occur in the short ITI/dynamic, but not fixed, tone condition corroborate this assumption, as well as results from other Pavlovian conditioning experiments utilizing dynamic cues in a head-fixed setup(11, 19). In contrast, experiments involving freely moving animals do not require explicitly dynamic cues because the sensory feedback from navigating though the environment presumably functions in the same way to indicate proximity to reward. Future experiments can investigate this further by characterizing the role of sensory feedback indicating reward proximity in mediating dopamine ramps. One set of observations superficially inconsistent with our assumption of the necessity of a sequence of external cues is that ramping dopamine dynamics can be observed even when only internal states signal reward proximity (e.g., timing a delayed action)(10, 13, 14). In these cases, however, animals were required to actively keep track of the passage of time, which therefore strengthens an internal progression of neural states signaling temporal proximity to reward. We speculate that once learned, these internal states could serve the role of external cues in the ANCCR framework. Previously, we argued against this kind of assumption in learning theories(46). Our earlier position was that it is problematic to assume fixed internal states that pre-exist and provide a scaffold for learning, such as in temporal difference learning. This is because these pre-existing states would need to already incorporate information that can only be acquired during the course of learning(46). Unlike this position, here we are merely speculating that after learning, an internal progression of states can serve the function of externally signaled events. Similarly, we have previously postulated that such an internal state exists during omission of a predicted reward, but only after learning of the cue-reward association(32).

Though the experiments in this study were motivated by the ANCCR framework, they were not conducted to discriminate between theories. As such, it is also possible to rationalize these results in the context of other models of dopamine ramps. In the value model, dopamine is thought to represent the discounted sum of future rewards(4–6, 18). The shape of this value function, and thus the predicted dopamine dynamics, is determined by the discount factor, γ. If there is a low γ (i.e., greater discounting), then the corresponding value function produces a steeper ramp. Consequently, it is possible to use the value model to explain our results if one assumes that a shorter ITI causes greater temporal discounting. The basis for such an assumption is unclear, though it has been suggested that the overall temporal discounting in an environment depends on reward rates(47, 48). Furthermore, we do not find evidence for dopamine ramps acting as a value signal to directly increase motivation. This is because we find similar trial-related behaviors in conditions with and without dopamine ramps. Thus, ITI dependent emergence of dopamine ramps for the same trial parameters provides strong constraints for the motivational role of dopamine ramps.

Substantial efforts have been made to account for the phenomenon of dopamine ramps as a temporal difference RPE(11, 18–20, 36). As with the value model, simulated dopamine responses in temporal difference RPE models are also modulated by the discount factor, γ. It has been proposed that temporal discounting in the dopamine system depends on the cue-reward delay(35). In our experiment, however, the cue-reward delay is not the key variable determining the presence of ramps; instead, it is the ITI. Another work has also proposed that a spectrum of discount factors can explain diverse activity profiles of single dopamine neurons(36). Specifically, monotonic upward ramps were simulated using a high γ (i.e., weaker discounting). Therefore, in this model, one would need to assume that shorter ITIs cause weaker temporal discounting to produce steeper ramps. Notably, this is in the opposite direction as the value model. Overall, it is unclear whether any fundamental principle predicts an ITI dependent change in temporal discounting in the dopamine system to allow RPE to explain our results. Similarly, whether other models of dopamine ramps(12, 18, 33) can capture an ITI dependent emergence of dopamine ramps remains to be explored.

While it is thus possible to rationalize our results using alternative theories of dopamine, ANCCR provides a principled and parsimonious explanation. Given that the foundation of ANCCR is looking back in time for causes of rewards, it is clear that differences in memory maintenance via eligibility traces will have profound implications on predicted dopamine signaling. For example, when the ITI is short and rewards are being frequently delivered, it intuitively makes sense that the eligibility trace time constant needs to be small; this is because the time window over which one would want to search for potentially causal cues is going to be shorter in this situation. We formalize this intuition by postulating that the eligibility trace time constant adapts to the overall event rates for efficient coding (Supplementary Note 1). In the case of a dynamic progression of cues, the cues closer in time to the reward will have higher causal power, and thus higher ANCCR, resulting in a dopamine ramp.

Our ANCCR simulations motivated the experiments, but we did not explicitly intend to fit the data. Accordingly, there are several details of the experiments that we did not include in the simulations. First, animals were trained initially using a long ITI (Pavlovian) or medium ITI (VR). This may explain a discrepancy between the simulations and experimental results: the cue onset response in the short ITI condition is small but positive in the experiment but negative in ANCCR. This discrepancy may be because the cue onset was already learned to be meaningful prior to the short ITI condition, thereby resulting in a stronger cue onset response in the experimental data. Further, we did not explicitly model potential trial-by-trial changes in eligibility trace time constant, sensory noise, internal threshold, local mechanisms controlling dopamine release, or sensor dynamics. Thus, we did not expect to capture all experimental observations in the motivating simulations. Regardless of such considerations, the current results provide a clear constraint for dopamine theories and demonstrate that an underappreciated experimental variable determines the emergence of mesolimbic dopamine ramps.